---
title: "Using simulated data for FISH 572"
author: "FISH 572, Winter 2026"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
library(ggplot2)
```

# Introduction
The mid-course project for Fish 572 is to analyze survey data, which may be a data set of their choice or simulated data provided by the instructors. This document describes how to use one of the simulated datasets. The methods described here may be extended to a real survey dataset.

# Simulated data
Data were simulated across a square grid of 100 X 100 cells with depth contours. Here is a plot of the locations on the grid with deeper depths shown in darker blue.

```{r grid}
# make sure in the correct directory
dat <- readRDS("coursework/simulations/sim_data/sim_dat_1.RDS")
predictor_dat <- readRDS("coursework/simulations/sim_data/grid.RDS")
predictor_dat$negDepth <- -1 * predictor_dat$depth

ggplot(dplyr::filter(predictor_dat, year == 1), aes(X, Y)) +
  geom_tile(aes(fill = negDepth),colour=rgb(0,0,0,0.1),linewidth=0.2) +
  scale_color_gradient2() + 
  scale_x_continuous(expand = c(0, 0)) + # Remove expansion on x-axis
  scale_y_continuous(expand = c(0, 0)) # Remove expansion on x-axis
```


There are eight (8) columns in a simulated dataset:
- year: the year, being 1:6;
- X: the location along the x-axis;
- Y: the location along the y-axis;
- eta: the simulated values;
- observed: the observed value of the simulated data (i.e. observation error applied to eta);
- eta_scaled: the simulated data scaled by the mean of simulated data in a reference simulated dataset;
- observed_scaled: the observed data scaled by the mean of observations in a reference simulated dataset;
- depth_scaled: the depth scaled by subtracting the mean and dividing by the standard deviation.



```{r simulated}
#plot eta
ggplot(dplyr::filter(dat, year == 1), aes(X, Y)) +
  geom_tile(aes(fill = eta),colour=rgb(1,1,1,0.05),linewidth=0.2) +
  scale_color_gradient2() + 
  scale_x_continuous(expand = c(0, 0)) + # Remove expansion on x-axis
  scale_y_continuous(expand = c(0, 0)) # Remove expansion on x-axis
```
```{r observed}
#plot observed
ggplot(dplyr::filter(dat, year == 1), aes(X, Y)) +
  geom_tile(aes(fill = observed),colour=rgb(1,1,1,0.05),linewidth=0.2) +
  scale_color_gradient2() + 
  scale_x_continuous(expand = c(0, 0)) + # Remove expansion on x-axis
  scale_y_continuous(expand = c(0, 0)) # Remove expansion on x-axis
```

# Sampling from the simulated data
One potential use of simulated data is to examine alternative sampling designs. In this example, we sample from the 100 X 100 grid in year 1 using simple random sampling (SRS).
```{r SRS}
source("coursework/simulations/functions_simulation.R")
set.seed(245)
n <- 1500
year <- 1
dat.yr <- dplyr::filter(dat, year == 1)
sampleRowNum <- sample(1:nrow(dat.yr),n,replace=F)
sampleDat <- dat.yr[sampleRowNum,]
# create sample (as if you were a boat sampling once in each sampled grid cell)
obsCV <- 0.2  # if using "observed" set this to zero (observation error already applied)
catchabilityPars <- list(mean=1,var=0.1) #related to gear selectivity (and availability to net)
theSample <- sampleGrid.fn(sampleDat, obsCV, catchabilityPars,varName="eta")


#plot sample
ggplot(theSample, aes(X, Y)) +
  geom_tile(aes(fill = observation)) +
  scale_color_gradient2() + 
  scale_x_continuous(expand = c(0, 0)) + # Remove expansion on x-axis
  scale_y_continuous(expand = c(0, 0)) # Remove expansion on x-axis
```

# Analysis
Now that you have your samples, the goal is to estimate the population in all grid cells across the entire sampling area (100 X 100 grid for the simulated data). In other words, you want to create the simulated map but you have not sampled all cells, and those that you did sample are subject to uncertainty. There are numerous ways to analyze the data, which can be lumped into ``design-based'' and ``modelled''. 



<!-- \pagebreak -->
